---
title: "Writing Functions"
description: |
  This is a tutorial for how to write a function in R.
author:
  - name: Sarah Dimakis
date: 05-13-2021
output:
  distill::distill_article:
    toc: true
    toc_float: true
    self_contained: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load libraries and read in data

```{r}

library(rio)
library(tidyverse)
library(kableExtra)
data <- import(here::here("data", "nhanes_1999-2016.csv")) %>% 
  mutate(food_security = case_when(hh_food_secure == 1 ~ "Fully food secure",
            hh_food_secure == 2 ~ "Marginally food secure",
            hh_food_secure == 3 ~ "Food insecure without hunger",
            hh_food_secure == 4 ~ "Food insecure with hunger"))

```

# Functions in R

Let's first create a function called `my_pet()` that will print out a statement about your pet. For best practice, you should try to name your function something descriptive. It should also not be named another function that is popular like `mean` because it will overwrite the default `mean` function for your script.

```{r}
my_pet <- function(pronoun, animal, verb){
  paste0(str_to_title(pronoun), " is a ", animal, " who likes to ", verb, ".")
}

```

The **arguments** of the function (which are called `formals`) are what the user supplies the function to get their desired output. You can see the formals of a function in R by using the `formals()` function. Now, in order to use the function, you can supply it with your desired arguments.



```{r}
my_pet("she", "dog", "play outside")
my_pet("he", "cat", "sleep")
my_pet("she", "lizard", "eat")

```

If you aren't sure what the `formals()` of a function are, you can access them with the `formals()` function.

```{r}
formals(my_pet)
```

The **body** of the function is where the function takes in the `formals` and creates the output. You can see the body of a function in R by using the `body()` function.

```{r}
body(my_pet)
```


# Warnings

If someone else (or your future self) is going to use your function, it is helpful to embed warnings (via `warning()`) or errors (via `stop()`) into your code to explain why the code will not work (or if it will not work as expected). In this example, I am also demonstrating the structure of an if else statement. I am saying that if each argument is a character, then I can proceed. Otherwise, I should produce a warning. 


```{r}
my_pet <- function(pronoun, animal, verb){
  if(is.character(pronoun) & is.character(animal) & is.character(verb)){
  paste0(str_to_title(pronoun), " is a ", animal, " who likes to ", verb, ".")}
  else{
    warning("You need to input the pronoun, animal, and verb as a character variable. Returning Null.")
    return()
  }
}

my_pet("she", "dog", "play outside")
my_pet("she", 3, "play outside")


```


# Example function 1

What makes a particularly good function is code that you write a lot. Additionally, you want to make a function that is simple and only does one thing. For this first function, I am going to create a function that will create a "total" column that calculates the total observations across a grouping variable.

Before I create a function, I first like to try to make it work in a single case.

```{r}
a <- count(data, year, gender)
a

b <- count(data, year)
b

b <- rename(b, "total" = "n")
b

left_join(a,b)
```

```{r}
total_grouping <- function(data, var, grouping){
  v1 <- enquo(var) 
  v2 <- enquo(grouping)
  
  a <- count(data, !!v1, !!v2)
  b <- count(data, !!v1)
  b <- rename(b, "total" = "n")
  c <- left_join(a,b)
  na.omit(c)
}
```

```{r}
total_grouping(data, year, gender)
```

Now I can do it with other variables!

```{r}
total_grouping(data, year, race_ethnic) %>% 
  head(10)

total_grouping(data, year, food_security) %>% 
  head(10)
```

# Example function 2

You can use a previous function you have defined earlier in a script in a new function! Here, I am extending the previous function to make a column that will calculate the percentage of the count in a given year. 

Let's first try an example.

```{r}
total_grouping(data, year, gender) %>% 
  mutate(percent = n/total * 100,
         percent = round(percent, 2),
         percent = paste0(percent, "%"))

```

Okay, now we're ready to generalize.
```{r}
percent_grouping <- function(data, var, grouping){
  v1 <- enquo(var) 
  v2 <- enquo(grouping)
  
  total_grouping(data, !!v1, !!v2) %>% 
  mutate(percent = n/total * 100,
         percent = round(percent, 2),
         percent = paste0(percent, "%")) 
}


```

```{r}
total_grouping(data, year, gender)

total_grouping(data, year, race_ethnic) %>% 
  head(10)

total_grouping(data, year, food_security) %>% 
  head(10)

```

# Tables

Now, let's make a table with our output.

```{r}
temp_table <- percent_grouping(data, year, gender) %>% 
  select(year, gender, percent) %>% 
  pivot_wider(names_from = gender,
              values_from = percent) %>% t(.)
table <- temp_table[2:3,]
colnames(table) <- temp_table[1,] 
table %>% kbl() %>% 
  kable_styling(bootstrap_options = "striped", full_width = F) %>% 
  row_spec(0, angle = -45)


```


I can generalize this with a function and I'll make tables for other variables! 

```{r}
my_table <- function(data, var, grouping){
  v1 <- enquo(var) 
  v2 <- enquo(grouping)
  temp_table <- percent_grouping(data, !!v1, !!v2) %>% 
    select(!!v1, !!v2, percent) %>% 
    pivot_wider(names_from = !!v2,
                values_from = percent) %>% t(.)
  table <- temp_table[2:nrow(temp_table),]
  colnames(table) <- temp_table[1,] 
  table %>% kbl() %>% 
    kable_styling(bootstrap_options = "striped", full_width = F) %>% 
    # row_spec(0, angle = -45, hline_after = TRUE)
    row_spec(0, hline_after = TRUE)
}

```

```{r}
my_table(data, year, gender)

my_table(data, year, race_ethnic)

my_table(data, year, food_security)
```

