---
title: "Writing Functions"
description: |
  This is a tutorial for how to write your first custom function in R.
author:
  - name: Sarah Dimakis
date: 05-13-2021
categories:
  - custom functions
output:
  distill::distill_article:
    toc: true
    toc_float: true
    self_contained: false
editor_options: 
  chunk_output_type: console
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load libraries and read in data

For this tutorial, you will first need to load `tidyverse` for data wrangling and `kableExtra` for tables. Next, I am using `rio::import` to import the cleaned data file to my environment. Note, you may need to change the file path if your file structure is not the same as mine. Last, I am using `mutate` and `case_when` to label the levels of the food_secure variable.


```{r}

library(tidyverse)
library(kableExtra)

data <- rio::import(here::here("data", "nhanes_1999-2016.csv")) %>% 
  mutate(food_security = case_when(hh_food_secure == 1 ~ "Fully food secure",
            hh_food_secure == 2 ~ "Marginally food secure",
            hh_food_secure == 3 ~ "Food insecure without hunger",
            hh_food_secure == 4 ~ "Food insecure with hunger"))

```

# Functions

A **function** is code that carries out an operation. For example `+` is a function that carries out the operation addition.

```{r}
2 + 3
```
In algebra, you may recall learning functions such as $f(x,y) = x^2+y$, where you put in inputs `x = 2` and `y = 1`, the function computes the operation, $2^2+1$, and outputs $5$. Similarly, when using a function in R, it takes a form like `f(x,y,z..)` where f is the function name and x,y,z... are the arguments of the function.

```{r}
`+`(2,3)
```

Let's create our first function in R called `my_pet()` that will print out a statement about your pet. For best practice, you should try to name your function something descriptive. It should also not be named another function that is popular like `mean` because it will overwrite the default `mean` function for your script.

```{r}
my_pet <- function(pronoun, animal, verb){
  paste0(str_to_title(pronoun), " is a ", animal, " who likes to ", verb, ".")
}

```





The **arguments** of the function (which are called `formals`) are what the user supplies the function to get their desired output. You can see the formals of a function in R by using the `formals()` function. 

```{r}
formals(my_pet)
```

The **body** of the function is where the function takes in the `formals` and creates the output. You can see the body of a function in R by using the `body()` function.

```{r}
body(my_pet)
```

Now, in order to use the function, you can supply it with your desired formals.

```{r}
my_pet(pronoun = "she", 
       animal = "dog", 
       verb = "play outside")


```
When you use your function, you can drop the argument names as long as you keep the same order.

```{r}
my_pet("he", "cat", "sleep")
```

If you want to use arguments out of the order they were defined in, you will need to label them.

```{r}
my_pet(animal = "lizard",
       pronoun ="she", 
       verb = "eat")

```
# Default settings

You can set a "default" setting for an argument. This is the setting that occurs when the user does not specify the argument. 

```{r}
my_pet2 <- function(pronoun, animal, verb = "dance"){
  paste0(str_to_title(pronoun), " is a ", animal, " who likes to ", verb, ".")
}
```

```{r}
my_pet2(pronoun = "she",
        animal = "dog")
```

The user can overwrite the default.

```{r}
my_pet2(pronoun = "she",
        animal = "dog",
        verb = "cuddle")
```


# Errors and warnings

If someone else (or your future self) is going to use your function, it is helpful to embed errors with `stop()` and/or warnings with `warning()`  into your code to explain why the code will not work (or if it will not work as expected). 


```{r}
# I am using an if else structure
# If the user inputs "cat" for animal, the function will 
# throw an error and say "Sorry, this function doesn't work # for people who own cats"
# If they use input any other animal, it will work

my_pet3 <- function(pronoun, animal, verb = "stretch"){
  if(animal == "cat"){
    stop("Sorry, this function doesn't work for people who own cats.")
  }
  else{
    paste0(str_to_title(pronoun), " is a ", animal, " who likes to ", verb, ".")
    }
}

```

```{r, error = TRUE}
# This works
my_pet3("she", "dog")

# This throws an error
my_pet3("he", "cat")

```

Instead of an error, you might just want to throw a warning, but still allow the function to work. 

```{r}

my_pet4 <- function(pronoun, animal, verb = "stretch"){
  if(animal != "dog"){
    warning(paste0("Really? A ", animal, "? You should really consider getting a dog."))
  }
  paste0(str_to_title(pronoun), " is a ", animal, " who likes to ", verb, ".")
    
}

```


```{r, warning= TRUE}

my_pet4("he", "cat")
my_pet4("he", "fish", "swim")


# works as normal
my_pet4("she", "dog")

```


These are silly examples, but there are many reasons why you would want the function to output a warning or error. For example, you may want the function to warn someone if their input is not going to give them the output they expect, or if the function will not work for certain input. This will help the user work with your function to get their desired output.

# Example function 1

What makes a particularly good function is code that you write a lot. Additionally, you want to make a function that is simple and only does one thing. For this first function, I am going to create a function that will create a "total" column that calculates the total observations across all levels of a grouping variable.

Before I create a function, I first like to try to make it work in a single case.

```{r}
a <- count(data, year, gender)
a

b <- count(data, year)
b

# Join a and b by year
# There are two n's so I am also changing the suffix of n so that they are labeled more clearly

left_join(a,b, by = "year", 
          suffix = c("_group", "_total"))
```

Now, let's try to generalize it.


```{r, error = TRUE}
total_grouping <- function(data, var, grouping){
  a <- count(data, var, grouping)
  b <- count(data, var)
  c <- left_join(a,b,by = "year", 
          suffix = c("_group", "_total"))
}
total_grouping(data, year, gender)

```

Unfortunately, from running this code, we get an error that says that the columns `var` and `grouping` are not found. This can happen when you use tidyverse functions to write functions because the tidyverse uses what is called **non-standard evaluation**. NSE makes the functions really user friendly. For example, when you use dplyr functions like `select(data, year)` or `group_by(data, year)`, the function knows that year is referring to `data$year` and not `year` from your global environment. However, this causes trouble when you try to use `select()` or `group_by()` (or in our case, `counts()`) in your own function. R looks for the column names (in our case, `var` and `grouping`) in your global environment but can't find them. In order to  work around this, we need to also use NSE. Here are two ways you can write the function with NSE:

1. {{}} Syntax

```{r}
total_grouping <- function(data, var, grouping){
  a <- count(data, {{var}}, {{grouping}}) ##wrap the variables in {{}}
  b <- count(data, {{var}})
  left_join(a,b,by = "year", 
          suffix = c("_group", "_total"))
}

total_grouping(data, year, gender) %>% head()
```

2. Quote the variables

```{r}
total_grouping <- function(data, var, grouping){
  v1 <- enquo(var)  #quote the variables
  v2 <- enquo(grouping)
  
  a <- count(data, !!v1, !!v2) # use !! to evaluate the quoted variables
  b <- count(data, !!v1)
  left_join(a,b,by = "year", 
          suffix = c("_group", "_total"))
}

total_grouping(data, year, gender) %>% head()

```

The great thing about functions is that we can now use our function with other variables, without having to copy,paste, and change the numbers. This cuts down on mistakes and makes your code easier to read.


```{r}
total_grouping(data, year, race_ethnic) %>% 
  head()

total_grouping(data, year, food_security) %>% 
  head()
```

# Example function 2

You can use a previous function you have defined earlier in a script in a new function! Here, I am extending the previous function to make a column that will calculate the percentage of the count in a given year. 

Let's first try an example.

```{r}
total_grouping(data, year, gender) %>% 
  mutate(percent = n_group/n_total * 100, #make a new variable called percent 
         percent = round(percent, 2),
         percent = paste0(percent, "%")) #round it and add a % sign

```

Okay, now we're ready to generalize. Note that I'm using NSE here too.

```{r}
percent_grouping <- function(data, var, grouping){
  total_grouping(data, {{var}}, {{grouping}}) %>% 
  mutate(percent = n_group/n_total * 100,
         percent = round(percent, 2),
         percent = paste0(percent, "%")) 
}


```

```{r}
percent_grouping(data, year, gender) %>% head()

percent_grouping(data, year, race_ethnic) %>% head()


```

# Tables

Now, let's make a table with our output. These tables will tell us how the demographics of our sample changed from year to year.

```{r}
# creates an ugly first draft 
temp_table <- percent_grouping(data, year, gender) %>% 
  select(year, gender, percent) %>% 
  pivot_wider(names_from = gender,
              values_from = percent) %>% t(.)

temp_table

# moves the first row to the title
table <- temp_table[2:3,]
colnames(table) <- temp_table[1,] 
rownames(table) <- rownames(table) %>% str_to_title()

#stylized table
table %>% kbl() %>% 
  kable_styling(bootstrap_options = "striped", full_width = F)  %>% #gives me a stylized striped table
  row_spec(0, angle = -45) #rotates column names


```


I can generalize this with a function so I can make tables for other variables! 

```{r}
my_table <- function(data, var, grouping){
  temp_table <- percent_grouping(data, {{var}}, {{grouping}}) %>% 
    select({{var}}, {{grouping}}, percent) %>% 
    pivot_wider(names_from = {{grouping}},
                values_from = percent) %>% t(.)
  
  table <- temp_table[2:nrow(temp_table),]
  colnames(table) <- temp_table[1,] 
  rownames(table) <- rownames(table) %>% str_to_title()
  
  table %>% kbl() %>% 
    kable_styling(bootstrap_options = "striped", full_width = F) %>% 
    row_spec(0, angle = -45)
}

```

```{r}
my_table(data, year, gender)

my_table(data, year, race_ethnic)

my_table(data, year, food_security)
```

Over time, as you use your function with different examples, you may want to tweak it. For example, I noticed by using the food security variable that I never explictly removed NAs. You may want to always remove NAs. Or, you can give your user an option to remove NAs or not by creating an argument called "remove.nas" and set it to TRUE to remove NAs by default. Then, if the user wants to see the NAs, they can set it to FALSE.

Additionally, I noticed that I made sure that my row names were capitalized, but I didn't do that for the column names because it wasn't relevant here (in all my examples the column names were numbers). This is why it is helpful to test your functions under as many different conditions as you can think of!



