---
title: "Data cleaning with {purrr}"
description: |
  A tutorial for using {purrr} for data cleaning.
author:
  - name: Lea Frank
categories:
  - map
  - map_df
  - map2
output:
  distill::distill_article:
    toc: true
    toc_float: true
    self_contained: false
    css: style.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(fs)
library(knitr)

# devtools::install_github("gadenbuie/xaringanExtra")
xaringanExtra::use_clipboard()
```

The `{purrr}` package provides a number of helpful functions for loading in multiple data files and iterating data processing steps over multiple data frames. This tutorial will provide step-by-step instructions that how just how powerful the `map` family of functions can be. 

## Batch Loading

The original data files downloaded from the *NHANES* website were separated by year and data type (i.e. demographics and FSQ), resulting in a number of data files. Rather than loading in each file one at a time, the `purrr::map` family of functions provides powerful tools for loading in large amounts of data. 

### Create a list with the file paths

The `here::here` function is first used to define the path to the data files. 

```{r}
(rootpath <- here::here("data"))
```

The `dir_ls` function from the `{fs}` package creates a vector of the paths for the files listed within `rootpath`.

```{r}
fnames <- fs::dir_ls(rootpath)
head(fnames)
```

Importantly, this vector is named with the associated paths. That is, each element of the vector is attributed a name that is also a character string of the file path. I know this sounds like redundant information, but it actually comes in handy when using `map_df` as shown below.
 
```{r}
names(head(fnames))
```

The demographics and FSQ will be loaded separately since they will require different data cleaning steps. Regular expression can be used with the `fs::dir_ls` function to filter files that match a specific pattern.

```{r}
# list file paths for files that contain the pattern "DEMO"
demo_paths <- dir_ls(rootpath, regexp = "DEMO")

# list file paths for files that contain the pattern "FSQ"
fsq_paths <- dir_ls(rootpath, regexp = "FSQ")
```

### Use `purrr::map` to read files

The `purrr::map` family of functions allows us to iterate a function over elements of a list or vector. In this case, we want to use `rio::import` to read in the data files using the vectors of file paths. 

There are two ways in which we can read in the data. The first option is to use the `purrr::map` function to read each data frame into a separate element of a list. This can be useful if you want to use `map` to iteration the same functions, like data cleaning, over each data frame.  

```{r}
# Loading the data files into separate lists
demos_list <- map(demo_paths, rio::import)
fsq_list <- map(demo_paths, rio::import)
```

Another option is to use `purrr::map_df` to read each data file into a single data frame. Importantly, this function allows you to specify the `.id`, which will create a variable to identify each iteration of the function. As I mentioned earlier, the `fs::dir_ls` function will assign names to each path. Using `map_df` on the named list of paths will generate a column in the data frame output that assigns the path name to each iteration of `rio::import`. 

```{r, layout="l-body-outset"}

# Loading the data files into a single data frame
demos <- map_dfr(demo_paths, rio::import, .id = "file")
kable(head(demos[,1:4]))
fsq <- map_dfr(fsq_paths, rio::import, .id = "file")
kable(head(fsq[,1:4]))

```

While both options are valid, which one you choose will ultimately depend on what you intend to do with the output. If I wanted to keep the data frames separated by year so I could more easily process and analyze the data, `map` would might be the better approach. 

If I wanted to create a single data file that contains survey data from all years (1999-2016) that could then be shared with collaborators, I might choose to use `map_dfr` to create a single data frame. This option is demonstrated below in **My Final Solution**. 

Let's take a look at how `map` can be used to iterate data cleaning steps over a list of data frames.

## Data cleaning 

Let's start with the list of data frames for the demographics surveys (`demos_list`). Each element of the list contains the demographics survey data for a specific year range (e.g. 1999-2000 or 2013-2014). If we want to apply the same data cleaning steps to each data frame, we can use `map` to iterate those data cleaning functions to each element of the list. 

First, let's start by creating a variable in each data frame for the `year`. For this step, we can use `map2` to iterate through the file paths (`demo_paths`) and list of data frames (`demos_list`) in parallel. The year will be extracted from the file path and applied to the corresponding data frame. 

This time, rather than using an existing function, we can create our own. When writing a function to use within the map functions, it must be wrapped within the curly brackets `~{ [some custom function] }`. To call each input, you would use `.x` (for the first input, `demo_paths`) and `.y` (for the second input, `demos_list`).

```{r, }
demos_list_clean <- map2(demo_paths, demos_list, ~{
  # create a new variable in demos_list for th year
  .y %>% 
    mutate(year = str_extract(.x, "\\d{4}-\\d{4}")) 
                  # extracts the pattern that matches: 4 digits-4 digits (e.g. 2001-2002)
})

kable(head(demos_list_clean[[1]]["year"]))
```

Next, I want to select the variables of interest and rename them to something more clear. This time, I will need to return the data frame in order for it to save in the output.

```{r, layout="l-body-outset"}
demos_list_clean <- map(demos_list_clean, ~{
  # selecting variables of interest
  .x <- .x %>% 
    select(year, SEQN, RIDAGEYR, RIAGENDR, 
         RIDRETH1, DMDEDUC2, DMDEDUC3)
  
  # renaming the columns
  names(.x) <- c("year", "id","age","gender",
                  "race_ethnic","educ_adult","educ_child")
  
  # return the new .x
  return(.x)
})

kable(head(demos_list_clean[[1]]))
```

Finally, I may want to convert some of the categorical variables into factors and add labels to the different levels.

```{r, layout="l-body-outset"}
demos_list_clean <- map(demos_list_clean, ~{
  
  # create new factors for gender and race/ethnicity
  .x %>% 
    mutate(gender = factor(gender, labels = c("male","female")),
           race_ethnic = factor(race_ethnic, labels = c("mexican-american",
                                                        "other-hispanic",
                                                        "non-hispanic-white",
                                                        "non-hispanic-black",
                                                        "other-race")))
})

kable(head(demos_list_clean[[1]]))
```

These steps can all be combined to produce the following code:

```{r}
demos_list_clean <- map2(demo_paths, demos_list, ~{
  .y <- .y %>% 
    mutate(year = str_extract(.x, "\\d{4}-\\d{4}"),
           RIAGENDR = factor(RIAGENDR, labels = c("male","female")),
           RIDRETH1 = factor(RIDRETH1, labels = c("mexican-american",
                                                        "other-hispanic",
                                                        "non-hispanic-white",
                                                        "non-hispanic-black",
                                                        "other-race"))) %>% 
    select(year, SEQN, RIDAGEYR, RIAGENDR, 
         RIDRETH1, DMDEDUC2, DMDEDUC3)
  names(.y) <- c("year", "id","age","gender",
                  "race_ethnic","educ_adult","educ_child")
  return(.y)
})

```

## My final solution

As I mentioned earlier, another possible route was to read the data files into a single data frame. Since the output was intended to be shared with collaborators on this blog, I decided this was the optimal solution. The same data cleaning steps conducted above can then be applied to the single data frame.

```{r}

demos <- dir_ls(rootpath, 
                 regexp = "DEMO") %>% # listing demographics files
  # batch loading demographics files
  map_dfr(rio::import, .id = "file") %>% 
                       # .id = "file" creates a variable with the file name
  # creating a variable for year from the file name id
  mutate(year = str_extract(file, "\\d{4}-\\d{4}"), 
         # creating factors for gender and race/ethnicity 
         RIAGENDR = factor(RIAGENDR, labels = c("male","female")),
         RIDRETH1 = factor(RIDRETH1, labels = c("mexican-american",
                                                "other-hispanic",
                                                "non-hispanic-white",
                                                "non-hispanic-black",
                                                "other-race"))) %>% 
  # selecting variables of interest
  select(year, SEQN, RIDAGEYR, RIAGENDR, 
         RIDRETH1, DMDEDUC2, DMDEDUC3)

# renaming the columns into something more legible
names(demos) <- c("year", "id","age","gender",
                  "race_ethnic","educ_adult","educ_child") 

fsq <- dir_ls(rootpath, 
                 regexp = "FSQ") %>% # listing questionnaire files
  # batch importing questionnaire files
  map_dfr(rio::import, .id = "file") %>% 
                       # .id = "file" creates a variable with the file name
  # creating new variables for year and food security
  mutate(year = str_extract(file, "\\d{4}-\\d{4}"), 
         hh_food_secure = ifelse(year == "1999-2000" | year == "2001-2002", 
                                 HHFDSEC, FSDHH)) %>% 
  # selecting variables of interest
  select(year, SEQN, hh_food_secure) 

# renaming the 'SEQN' column so that it can be joined with the demographics data
names(fsq)[2] <- "id" 
  
```

After creating data frames for the demographics and FSQ data, they were then joined by the `year` and `id`. 

```{r, layout="l-body-outset"}
df <- left_join(demos, fsq)
kable(head(df))
```

Finally, the data frame was saved into a single `.csv` file that could then be shared with collaborators. 

```{r}
write_csv(df, str_c(rootpath,"/nhanes_1999-2016.csv"))
```

